On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-12 Manuscript/manuscript.tex
Almost done with notes

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for this in the model through the batch
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
\eds{I was talking to my husband about this. Could the heat effect be justified 
to capture the variability in the amount of time between the ``set'' and the 
gun going off (assuming this is different from heat to heat)? E.g., quicker 
gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
but this effect would be shared across all athletes in the same heat.}
\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases when
including 2022.  The two effect model with venue
and heat effect was the best model, as indicated by an ANOVA test 
\eds{I think it may be better to call this a likelihood ratio test rather than
an ANOVA test, assuming that is the formal test being used} and decrease in
AIC.  For the entire data, the AIC for the venue effect only model was
$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, the two
effect model is superior and will be presented throughout the rest of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ &       & 0.048 & $-1.857$ & 0.152\\
      Heat Effect Only & $-2070.7$ & 0.073 &       & $-1.890$ & 0.135\\
      Full model       & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less than the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $<$ x = $10^{-3}$) & P(RxnTime $<$ x = $10^{-4}$) \\ 
   \midrule
   Old Pooled & $0.094$ & $0.084$ \\ 
   All Pooled & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

@@

Table~\ref{tab:Clusrankresults} shows the results generated from four tests in R.
We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
\texttt{B = 1000} indicating that the random permutation size was chosen to be 
1000, as discussed previously in Section~\ref{sec:rank}.  
%As the sample sizes are small, 65 and 49, we use \texttt{exact = TRUE} as
%that is preferred.  
In both cases, the national versus international comparison
produced stronger \eds{lower?} p values than the 2019 versus 2022 comparison, with the
national vs international \texttt{exact = TRUE} test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were a few months prior.  For the 2019 versus 2022
same athlete comparison with \texttt{exact = TRUE} argument, the p value was
significant at $\alpha = 0.01$, however when \texttt{exact = FALSE}, the p
value was higher and only significant at $\alpha = 0.05$. 
\eds{not sure the \texttt{exact = FALSE} results need to be reported here or in
Table 4 when we say we use \texttt{exact = TRUE} here and in the previous section.}
Nonetheless, these
tests help to show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

\begin{table}
  \centering
  \caption{}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10^{-2}$ \\
   2022 vs 2023 & $3.0\cdot10^{-3}$ & $1.0\cdot10^{-2}$ \\
   Nat vs Inter & $2.2\cdot10^{-16}$ & $1.0\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \caption{2019 vs 2022 compares data from the same athletes who competed at the
  2019 and 2022 World Track and Field Championships. 2022 vs 2023 compares data
  from the same athletes who competed at the 2022 and 2023 World Track and Field
  Championships. Nat vs Inter compares data from the same athletes who competed
  at the 2022 national-level championships and the 2022 World Track and Field
  Championships.}
  \label{tab:Clusrankresults}
\end{table}

@@

%Since 1985 Seiko Holding Corporations has served every World Athletics 
%Championships as the official timer \citep{wa2022seiko}.  Reaction times were
%not officially measured until 1999 in the men's 110 meter hurdles and women's
%100 meter hurdles. Seiko's technology for detecting 
%reaction times relies on the pressure that athletes exert on the plate when they 
%push off.  Their systems measure the time differential between when the ``gun" 
%goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
%If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
%as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
%Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
%the gun and it was luck that caused their abnormally high time.  At the World 
%Championships level, World Athletics want to remove that element of 
%luck and thus impose the 0.1 second barrier.
%\eds{I would argue that this paragraph should go in the discussion as there are 
%no results presented here}
%\of{I agree so I moved it here. I think it also could fit into the top of the
%data section as in section~\ref{sec:Data}. Thoughts?}

%\eds{I would suggest removing the last two sentences, as they would suggest your
% modeling with GLMM is not good}.  
%\of{I agree and they have been removed}

%If there are issues with reaction time such as athletes approaching the 0.1 
%second barrier again at the next World Championships (August 2023), then World 
%Athletics needs to adjust that threshold and allow for faster reaction times so 
%that athletes with superb reflexes are not penalized.  World Athletics did not
%do anything in response to the \citet{komi2009iaaf} study, but perhaps they 
%should now so that the story of Devon Allen is not repeated.

\section*{Appendix}
\label{sec:Appendix}
%Put downloadable data spreadsheet
The World Championships data are available from the World Athletics website:
\url{https://www.worldathletics.org/results/world-athletics-championships}.

\jy{You have countries beyond United States}
The United States Track and Field Championships data are available at: 
\url{https://www.flashresults.com/2022_Meets/Outdoor/06-23_USATF/}

\jy{Go over the references to see clues to clean in the bib source}
\bibliographystyle{chicago}
\bibliography{citations.bib}

>>>>

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.  Although 2022 is presented three times in
figure~\ref{fig:RankScatterplots}, the subset of athletes differs each time.
There may be athletes who competed in 2019 and 2022 and their national
championships data was available, but they did not compete in 2023, and thus their
times show up in the top two graphs.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for this in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
\eds{I was talking to my husband about this. Could the heat effect be justified 
to capture the variability in the amount of time between the ``set'' and the 
gun going off (assuming this is different from heat to heat)? E.g., quicker 
gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
but this effect would be shared across all athletes in the same heat.}
\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases when including 2022.  The two effect model with venue and heat
effect was the best model as evidenced by a decrease in the AIC compared to the
other models. For the entire data, the AIC for the venue effect only model was
$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, the two effect model is superior and will be presented throughout the rest
of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}
\of{Professor Yan recommended not even mentioning the likelihood ratio so
instead I will just talk about the AIC.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ & N/A & 0.048 & $-1.857$ & 0.152\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Full model       & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less than the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $< x = 10^{-3}$) & P(RxnTime $< x = 10^{-4}$) \\ 
   \midrule
   Excluding 2022 & $0.094$ & $0.084$ \\ 
   Including 2022 & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

@@

Table~\ref{tab:Clusrankresults} shows the results generated from four tests in R.
We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
\texttt{B = 1000} indicating that the random permutation size was chosen to be 
1000, as discussed previously in Section~\ref{sec:rank}.  
%As the sample sizes are small, 65 and 49, we use \texttt{exact = TRUE} as
%that is preferred.  
In both cases, the national versus international comparison
produced lower p values than the 2019 versus 2022 comparison, with the
national vs international \texttt{exact = TRUE} test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were a few months prior.  For the 2019 versus 2022
same athlete comparison with \texttt{exact = TRUE} argument, the p value was
significant at $\alpha = 0.01$, however when \texttt{exact = FALSE}, the p
value was higher and only significant at $\alpha = 0.05$. 
\eds{not sure the \texttt{exact = FALSE} results need to be reported here or in
Table 4 when we say we use \texttt{exact = TRUE} here and in the previous section.}
Nonetheless, these
tests help to show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

\begin{table}
  \centering
  \caption{}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10^{-2}$ \\
   2022 vs 2023 & $3.0\cdot10^{-3}$ & $1.0\cdot10^{-2}$ \\
   2022 National vs International & $2.2\cdot10^{-16}$ & $1.0\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \caption{2019 vs 2022 compares data from the same athletes who competed at the
  2019 and 2022 World Track and Field Championships. 2022 vs 2023 compares data
  from the same athletes who competed at the 2022 and 2023 World Track and Field
  Championships. Nat vs Inter compares data from the same athletes who competed
  at the 2022 national-level championships and the 2022 World Track and Field
  Championships.}
  \label{tab:Clusrankresults}
\end{table}

@@

\bibliographystyle{chicago}
\bibliography{citations.bib}
